Measuring directionality in double-beta decay and neutrino interactions with kiloton-scale scintillation detectors

### 2014-01-10 Direction_JINST_v8.tex
Added Christophs edits and committing Henrys referee response. --Lindley

The scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. The first step in the scintillation process is the transfer of energy from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns; more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator.

Due to the wavelength-dependence of the refractive index the speed of
light in the scintillator (see Equation (\ref{eqGroup})) increases
with increasing photon wavelengths for normal dispersion, with red
light traveling faster than blue light.  In order to study the
measurability of the time differences due to this chromatic dispersion, we simulated 5~MeV electrons at
the center of the sphere where we used instantaneous scintillation
emission with the quantum-efficiency applied, but not including a
transit-time spread. The higher energy provides larger photon statistics with which to evaluate the method, debug the simulation, and is of interest to neutrino-electron scattering experiments. The true hit time distributions of photoelectrons
were analyzed for scintillation light and Cherenkov light
separately. Photoelectrons coming from Cherenkov light are on average
created about 0.5~ns earlier than PEs from scintillation light. The
RMS values from PE time distributions for Cherenkov and scintillation
light are both about 0.5~ns. Note that these numbers include the
effect of the finite electron travel time.

The measurement of the arrival times of single photoelectrons is
affected by the transit-time spread (TTS) of the photodetectors, a
number which can be different by orders of magnitude depending on the
detector type. The default TTS of 0.1~ns ($\sigma$) can be achieved with large area picosecond photodetectors
(LAPPDs)\cite{Adams:2013nva,RSI_paper,PSEC4_paper,anode_paper} and possibly hybrid photodetectors
(HPDs)\cite{hpdThesis}; even significantly lower TTS numbers are
realistic with the LAPPD\cite{RSI_paper,PSEC4_paper,anode_paper}.

@@

\section{Conclusions}
The ability to reconstruct direction in kiloton-scale scintillation
detectors would be a major technological advance for neutrino
experiments, especially those also
searching for neutrino-less double-beta decay. More generally, this technique could
be applied wherever scintillation-based detectors are used. A \GEANT~simulation of a simple spherical detector corresponding to a kiloton
of scintillator shows that with timing on the order of 0.1~ns the directional Cherenkov light can be separated from the more abundant
scintillation light. The separation can be improved using
photodetectors with more red sensitivity and liquid scintillators with
a more narrow emission spectrum shifted to shorter
wavelengths. Furthermore, simple reconstruction algorithms adapted
from those for water Cherenkov detectors are able to reliably
reconstruct the position and direction of electrons with energies in the few MeV range.
More detailed simulation and advanced reconstruction algorithms will need
to be developed to move to more complicated event structure, such as those in neutrino-less double-beta decay. The technique already
appears promising and justifies future work on the topic.

>>>>

The scintillator-specific rise and decay times are the second effect that determines the timing in a scintillator detector. The first step in the scintillation process is the transfer of energy from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns; more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator.

Chromatic dispersion is the third effect that determines the timing in a scintillator detector. Due to the wavelength-dependence of the refractive index the speed of
light in the scintillator (see Equation (\ref{eqGroup})) increases
with increasing photon wavelengths for normal dispersion, with red
light traveling faster than blue light.  In order to study the time differences due to this chromatic dispersion, we used a simplified simulation of 5~MeV electrons at
the center of the sphere where we used instantaneous scintillation
emission with the quantum-efficiency applied, but not including a
transit-time spread. The higher energy 5~MeV electron provides larger photon statistics with which to evaluate the method, debug the simulation, and is of interest to neutrino-electron scattering experiments. The true hit time distributions of photoelectrons
were analyzed for scintillation light and Cherenkov light
separately. Photoelectrons coming from Cherenkov light are on average
created about 0.5~ns earlier than PEs from scintillation light. The
RMS values from PE time distributions for Cherenkov and scintillation
light are both about 0.5~ns. Note that these numbers include the
effect of the finite electron travel time.

The fourth effect determining the timing in a scintillator detector is the timing of the photodetectors. The measurement of the arrival times of single photoelectrons is
affected by the transit-time spread (TTS) of the photodetectors, a
number which can be different by orders of magnitude depending on the
detector type. The default TTS of 0.1~ns ($\sigma$) can be achieved with large area picosecond photodetectors
(LAPPDs)\cite{Adams:2013nva,RSI_paper,PSEC4_paper,anode_paper} and possibly hybrid photodetectors
(HPDs)\cite{hpdThesis}; even significantly lower TTS numbers are
realistic with the LAPPD\cite{RSI_paper,PSEC4_paper,anode_paper}.

@@

\section{Conclusions}

We developed a technique to separate scintillation and Cherenkov light to reconstruct direction of electrons with energies 5~MeV, 2.1~MeV and 1.4~MeV in a liquid scintillator detector. These energies were chosen to represent typical neutrino-electron elastic scattering energies and
$0\nu\beta\beta$ energies. The Cherenkov threshold for an electron in a typical liquid scintillator is $\sim$0.2~MeV

While scintillation light is isotropic, Cherenkov light with wavelengths above $\sim$370~nm carries the information about the direction of the electrons. All light shorter than $\sim$370~nm get absorbed and re-emitted isotropicaly as a part of the scintillation process. On average scintillation light is delayed with respect to the direct Cherenkov light due to chromatic dispersion and the finite rise time of the scintillation. Therefore early light contains directional information.

Using a GEANT~simulation of a spherical detector with radius of 6.5~m, we showed that for electrons originated at the center, Cherenkov light arrives on average $\sim$1~ns before the scintillation light. Chromatic dispersion alone accounts for $\sim$0.5~ns delay between scintillation and Cherenkov light. We then defined an early time window and found for a 5~MeV electron that the ratio of Cherenkov to scintillation light is 0.63 for photodetectors with TTS of 0.1~ns. This was degraded by a factor of 2.5 if current photodetectors with $\sim$1~ns resolution were used. This can be improved by 1.6 or 1.4  if more red sensitive photodetectors or scintillators with narrower emission spectra are used

We applied reconstruction algorithms from water Cherenkov detectors. QUANTATIVE STATEMENT ABOUT VERTEX DIRECTION PERFORMANCE.

We find the technique promising and plan to continue work on the topic. The ability to reconstruct direction in kiloton-scale scintillation detectors would expand capabilities of neutrino experiments, especially those also searching for neutrino-less double-beta decay. More generally, this technique could be applied wherever scintillation-based detectors are used.

%This work should include a more detailed simulation including second order effects such as reemission and a more detailed photodetector model including electronics thresholds. More advanced reconstruction methods are also needed to simultaneously fit position and direction and to handle the more complicated event structure found in neutrino-less double-beta decay.

%The separation can be improved using photodetectors with more red sensitivity and liquid scintillators with a more narrow emission spectrum shifted to shorter wavelengths. Furthermore, simple reconstruction algorithms adapted from those for water Cherenkov detectors are able to reliably reconstruct the position and direction of electrons with energies in the few MeV range. More detailed simulation and advanced reconstruction algorithms will need to be developed to move to more complicated event structure, such as those in neutrino-less double-beta decay.

### 2014-01-10 Direction_JINST_v9.tex
Added Christophs edits and committing Henrys referee response. --Lindley

Scintillation light is isotropic. At these low energies, it does not contain sufficient information to reconstruct the track of the outgoing particles, although at higher energies it may\cite{john}.  Cherenkov light is also
produced for electrons above threshold. Most is absorbed and re-emitted as part of the
scintillation processes; however some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle. The addition of directionality is a powerful tool
for background rejection, especially for $0\nu\beta\beta$ searches. In the high pressure TPCs, reduction factors on the order of $10^4$ have been achieved\cite{Gotthard}. It is also possible to look for new physics in the angular correlation of the emerging electrons\cite{newphysics0nuBB}, as has been proposed for the tracking-based detectors\cite{SuperNEMO}. The addition of a directional signal would make large-scale liquid scintillator detectors more competitive for the next generation of $0\nu\beta\beta$ searches and merits investigation.

@@

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. Figure \ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match a KamLAND-like
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene and 1.52~g/l PPO. The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}.  This is a standard scintillator. The most critical value is the attenuation length at 400~nm which is 25~m. The attenuation length drops precipitously at 370~nm from 6.5~m to 0.65~m at the cutoff wavelength at 360~nm.
Variations from the baseline KamLAND case are discussed below.

@@

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs2/PEtime_Cerenkov_scattered_vs_unscattered.pdf}
        \caption[]{Photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV). The default simulation is shown with optical scattering turned on: all photons (black), un-scattered photons (red dotted), and scattered (blue dashed). Scattering causes a very small tail at longer times as expected. \label{scattplot}}
        \end{center}
\end{figure}

@@

Four effects primarily contribute to the timing of the scintillator detector
system: the travel time of the particle, the time constants of the scintillation process, chromatic dispersion, and the timing of the photodetector.  First, the simulated travel time of a 5~MeV electron is 0.108$\pm$0.015~ns. This corresponds to an average path length of 3.1~cm and a final distance from the origin of 2.6~cm. The time until the electron drops below Cherenkov threshold is 0.106$\pm$0.015~ns. We note that due to scattering the final direction of the electron before it stops does not correspond to the initial direction; however before the electron drops below Cherenkov threshold the scattering angle is small and therefore the Cherenkov light encodes the direction of the primary electron. The scattering physics is handled by Geant4's ``Multiple Scattering" process which is valid down to 1~keV, where atomic shell structure becomes important\cite{geant4scatt}. In the energy range important for $0\nu\beta\beta$, a 1.4~MeV electron travels a total path length of 0.8~cm, has a distance from the origin of 0.6~cm in 0.030$\pm$0.004~ns  and takes 0.028$\pm$0.004~ns to drop below Cherenkov threshold. The scattering follows the same pattern.

@@

\begin{figure}[tbh]
        \begin{center}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq2_5p0.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_5p0.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq2_2p1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_2p1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq2_1p4.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_1p4.pdf}}
        \caption[]{\label{fig:reco} (Left) The reconstructed direction,
        $(p_x/|\vec{p}|, p_y/|\vec{p}|,
        p_z/|\vec{p}|)$, for the simulation of 1000 electrons. In the simulation the electrons are produced along the
        x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0), and originate
        from the center of the 6.5m-radius detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34.0~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Right) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation. From Top to Bottom, 5~MeV, 2.1~MeV and 1.4 Mev are shown.}
\end{center}
\end{figure}

@@

We developed a technique to separate scintillation and Cherenkov light to reconstruct direction of electrons with energies 5~MeV, 2.1~MeV and 1.4~MeV in a liquid scintillator detector. These energies were chosen to represent typical neutrino-electron elastic scattering energies and
$0\nu\beta\beta$ energies. The Cherenkov threshold for an electron in a typical liquid scintillator is $\sim$0.2~MeV

While scintillation light is isotropic, Cherenkov light with wavelengths above $\sim$370~nm carries the information about the direction of the electrons. All light shorter than $\sim$370~nm get absorbed and re-emitted isotropicaly as a part of the scintillation process. On average scintillation light is delayed with respect to the direct Cherenkov light due to chromatic dispersion and the finite rise time of the scintillation. Therefore early light contains directional information.

Using a GEANT~simulation of a spherical detector with radius of 6.5~m, we showed that for electrons originated at the center, Cherenkov light arrives on average $\sim$1~ns before the scintillation light. Chromatic dispersion alone accounts for $\sim$0.5~ns delay between scintillation and Cherenkov light. We then defined an early time window and found for a 5~MeV electron that the ratio of Cherenkov to scintillation light is 0.63 for photodetectors with TTS of 0.1~ns. This was degraded by a factor of 2.5 if current photodetectors with $\sim$1~ns resolution were used. This can be improved by 1.6 or 1.4  if more red sensitive photodetectors or scintillators with narrower emission spectra are used

We applied reconstruction algorithms from water Cherenkov detectors. QUANTATIVE STATEMENT ABOUT VERTEX DIRECTION PERFORMANCE.

We find the technique promising and plan to continue work on the topic. The ability to reconstruct direction in kiloton-scale scintillation detectors would expand capabilities of neutrino experiments, especially those also searching for neutrino-less double-beta decay. More generally, this technique could be applied wherever scintillation-based detectors are used.

>>>>

Scintillation light is isotropic. At these low energies, it does not contain sufficient information to reconstruct the track of the outgoing particles, although at higher energies it may\cite{john}.  Cherenkov light is also
produced for electrons above threshold. Most is absorbed and re-emitted as part of the
scintillation processes; however some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle. The addition of directionality is a powerful tool
for background rejection, especially for $0\nu\beta\beta$ searches. In the high pressure TPCs, reduction factors on the order of $\sim10^4$ have been achieved\cite{Gotthard}. It is also possible to look for new physics in the angular correlation of the emerging electrons\cite{newphysics0nuBB}, as has been proposed for the tracking-based detectors\cite{SuperNEMO}. The addition of a directional signal would make large-scale liquid scintillator detectors more competitive for the next generation of $0\nu\beta\beta$ searches and merits investigation.

@@

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. Figure \ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match a KamLAND-like
scintillator\cite{kamland2003}: 80\% n-dodecane, 20\% pseudocumene and 1.52~g/l PPO. The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length\cite{tajimaMaster} and
refractive index\cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 
0.13)\cite{tajimaThesis}, scintillator light yield (9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)\cite{ChrisThesis}.  This is a standard scintillator. The attenuation length at 400~nm, the position of the peak standard bialkali photocathod efficiency, is 25~m. The attenuation length drops precipitously at 370~nm from 6.5~m to 0.65~m at the cutoff wavelength at 360~nm.
Variations from the baseline KamLAND case are discussed below.

@@

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs2/PEtime_Cerenkov_scattered_vs_unscattered.pdf}
        \caption[]{Cherenkov photoelectron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV). The default simulation is shown with optical scattering turned on: all photons (black), un-scattered photons (red dotted), and scattered (blue dashed). Scattering causes a very small tail at longer times as expected. \label{scattplot}}
        \end{center}
\end{figure}

@@

Four effects primarily contribute to the timing of the scintillator detector
system: the travel time of the particle, the time constants of the scintillation process, chromatic dispersion, and the timing of the photodetector.  First, the simulated travel time of a 5~MeV electron is 0.108$\pm$0.015~ns. This corresponds to an average path length of 3.1~cm and a final distance from the origin of 2.6~cm. The time until the electron drops below Cherenkov threshold is 0.106$\pm$0.015~ns. We note that due to scattering the final direction of the electron before it stops does not correspond to the initial direction; however the scattering angle is small while the majority of Cherenkov light is produced. The Cherenkov light thus encodes the direction of the primary electron. The scattering physics is handled by Geant4's ``Multiple Scattering" process which is valid down to 1~keV, where atomic shell structure becomes important\cite{geant4scatt}. In the energy range important for $0\nu\beta\beta$, a 1.4~MeV electron travels a total path length of 0.8~cm, has a distance from the origin of 0.6~cm in 0.030$\pm$0.004~ns  and takes 0.028$\pm$0.004~ns to drop below Cherenkov threshold. The scattering follows the same pattern.

@@

\begin{figure}[tbh]
        \begin{center}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hDirZ_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxX_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxY_v1.pdf}}
        %%\subfigure{\includegraphics[scale=0.295]{graphs/hVtxZ_v1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq2_5p0.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_5p0.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq2_2p1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_2p1.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hDir_pq2_1p4.pdf}}
        \subfigure{\includegraphics[scale=0.375]{graphs2/hVtx_pq2_1p4.pdf}}
        \caption[]{\label{fig:reco} (Left) The reconstructed direction,
        $(p_x/|\vec{p}|, p_y/|\vec{p}|,
        p_z/|\vec{p}|)$, for the simulation of 1000 electrons. In the simulation the electrons are produced along the
        x-axis, $\vec{p}/|\vec{p}|$ = (1,0,0), and originate
        from the center of the 6.5m-radius detector, $\vec{r}$ =
        (0,0,0). Only photons with arrival time of $t<$ 34.0~ns are used
        in the reconstruction. The quantum efficiency of the bialkali
        photocathode is taken into account. (Right) The reconstructed
        vertex position, $(x,y,z)$, for the same simulation. From Top to Bottom, 5~MeV, 2.1~MeV and 1.4 MeV are shown.}
\end{center}
\end{figure}

@@

We have developed a technique to separate scintillation and Cherenkov light to reconstruct direction of electrons with energies 5~MeV, 2.1~MeV and 1.4~MeV in a liquid scintillator detector. These energies have been chosen to represent typical neutrino-electron elastic scattering energies and
$0\nu\beta\beta$ energies. The Cherenkov threshold for an electron in a typical liquid scintillator is $\sim$0.188~MeV

While scintillation light is isotropic, Cherenkov light with wavelengths above the absorption cutoff $\sim$370~nm carries the information about the direction of the electrons. All light shorter than this gets absorbed and re-emitted isotropically as a part of the scintillation process. On average scintillation light is delayed with respect to the direct Cherenkov light due to chromatic dispersion and and the finite time of the scintillation processes, therefore early light contains directional information.

Using a Geant4~simulation of a spherical detector with radius of 6.5~m and photodetectors with TTS of 0.1~ns, we have showed that for electrons originated at the center a time cut on the early light was effective at isolating the directional light, improving the ratio of Cherenkov light to scintillation light from 0.02 to 0.63. This ratio is degraded by a factor of 2.5 if current photodetectors with $\sim$1~ns resolution were used. This can be improved by factors of 1.6 or 1.4 if more red sensitive photodetectors or scintillators with narrower emission spectra are used.

Reconstruction algorithms developed for water Cherenkov detectors have been applied to this early light. The algorithms are able to converge on reasonable reconstructed vertices and directions for all simulated energies: 5~MeV, 2.1~MeV, 1.4~MeV. As expected, the reduction in photon statistics with lower energies leads to a broadening of the reconstructed vertex by 1.5~cm. Similarly, the mean reconstructed direction drops from XX to XX as the $\cos(\theta)$ distribution broadens.

We have shown that this technique promising, and we plan to continue work on the topic. The ability to reconstruct direction in kiloton-scale scintillation detectors would expand capabilities of neutrino experiments, especially those also searching for neutrino-less double-beta decay. More generally, this technique could be applied wherever scintillation-based detectors are used.
